RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-06 doc/paper.tex
Introduction: pocetak revidiranja

\author[1]{Aleksandar Kartelj}
\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
    \address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
          In this paper we solve the well-known, prominent problem of Symbolic regression, intensively studied over the last two decades. This problem have a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards  establishing  physical theory behind, thus helping establishing reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology,  modeling analytic representations of the exciton binding energy in the energy science, finding models for band gap of NaCl-type compounds
         in material science, just to name a few. We propose a new meta-heuristic based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are a several key concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used within the local search procedure to make it efficient and to fit the coefficients in models (solutions);  ($iii$) a novel fitness function is utilized which combines an accuracy score, error score, and the size of model into one, more compact score. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} to 14 other competitors from the literature considering different levels of noise in the input data.  Our method obtains new best exact percentages on the both benchmark sets comparing to the literature approaches. Moreover, new unbiased randomly generated set of problem instances are introduced where scalability and robustness of our method is thoroughly demonstrated in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

	The problem of symbolic regression (SR)~\cite{billard2002symbolic} has attracted many researchers over the last decade to study it intensively. SR can be seen as a generalization of the well known  concept of linear regression, i.e., polynomial regression~\cite{stimson1978interpreting}. All regression models in principle have the same task: given a set of a $n$-dimensional input data and the output data, the aim is to find a  mathematical expression (function) consisting of $n$ (input) variables that fits to the output data w.r.t. some in advance known measure.  This computationally intensive task is in general  provenly NP--hard~\cite{virgolin2022symbolic}. When aiming a model to be a linear combination of input variables, it represents the problem of linear regression. However, when there are some nonlinear relations between variables, which are relevant, linear regression models are not enough. This is the point where symbolic regression comes into the play. Unlike linear regression, it allows the search over the space of all possible mathematical formulas in order to find the best fitting ones able to  predict the output variable from the input variables. The base of constructing the explicit formula are the basis operations like addition and multiplication, as well as polynomial, trigonometric, exponential, and other elementary functions.  
	%Application: https://towardsdatascience.com/real-world-applications-of-symbolic-regression-2025d17b88ef

>>>>

\author[1]{Aleksandar Kartelj}
\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
    \address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
          In this paper, we solve the well-known, prominent problem of Symbolic regression, intensively studied within the last two decades. This problem has a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards establishing physical theory behind, thus helping establish reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modelling analytic representations of the exciton binding energy in energy science, and finding models for band gaps of NaCl-type compounds
          in material science, just to name a few. We propose a new meta-heuristic-based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are several key concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used within the local search procedure to make it efficient and to fit the coefficients in models (solutions);  ($iii$) a novel fitness function is utilized which combines an accuracy score, error score, and the size of the model into one, more compact score. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} to 14 other competitors from the literature considering different levels of noise in the input data.  Our method obtains new best exact percentages on both benchmark sets compared to the literature approaches. Moreover, a new unbiased randomly generated set of problem instances are introduced where the scalability and robustness of our method are thoroughly demonstrated in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

	The problem of symbolic regression (SR)~\cite{billard2002symbolic} has attracted many researchers over the last decade to study it intensively. SR can be seen as a generalization of the well-known concept of linear regression, i.e., polynomial regression~\cite{stimson1978interpreting}. All regression models in principle have the same task: given a set of $n$-dimensional input data and the output data, the aim is to find a  mathematical expression (function) consisting of $n$ (input) variables that fit the output data w.r.t. some in advance known measure.  This computationally intensive task is in general provenly NP--hard~\cite{virgolin2022symbolic}. When aiming a model to be a linear combination of input variables, it represents the problem of linear regression. However, when there are some nonlinear relations between variables, which are relevant, linear regression models are not enough. This is the point where symbolic regression comes into play. Unlike linear regression, it allows the search over the space of all possible mathematical formulas to find the best-fitting ones able to predict the output variable from the input variables. The basis of constructing the explicit formula is the basic operations like addition and multiplication, as well as polynomial, trigonometric, exponential, and other elementary functions.  \fxnote{Marko: došao dovde sa ovom sekcijom}
	%Application: https://towardsdatascience.com/real-world-applications-of-symbolic-regression-2025d17b88ef
